---
title: "Week 2 Lab"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the following exercise, we test the relationship between having a private tutor and their mathematics scores. 

```{r, include = FALSE}
library(tidyverse)
theme_set(theme_minimal())

n <- 10000 # how many samples we'll typically take
set.seed(12345)
# data frame containing noise variables, which we will then use to create each of the actual variables
ind <- data.frame(private_tutor_e = rnorm(n), 
                  reading_writing_e = rnorm(n), 
                  mathematics_e = rnorm(n))
```

## Create the data


```{r}
d <- data.frame(private_tutor_utility = -0.5 + ind$private_tutor_e) %>%
  mutate(private_tutor = ifelse(private_tutor_utility > 0, 1, 0),
         mathematics = 1 + private_tutor * 2 + ind$mathematics_e)
```

Note: The mechanism for deciding whether `private_tutor` should 
be 1 or 0 is the one presumed by probit models, but there's nothing particularly 
special about it. If you like to think about stochastic decision models then this 
is a trivial example of one (normally the mean, here -0.5, would be a function
of something more interesting, not just an intercept). In this style, 
think of `private_tutor_utility` 
as representing a large number of considerations influencing a 
person's decision to take on a tutor. Since these considerations are personal, 
and unobserved, they are represented by as one of the `_e` noise variables. 
But this is all computational convenience really.

OK, let's construct the *potential outcomes*
```{r}
d <- mutate(d,
            mathematics_0 = 1 + 0 * 2  + ind$mathematics_e,
            mathematics_1 = 1 + 1 * 2  + ind$mathematics_e)
```

Question: Are potential outcomes and the actual outcome independent of whether someone actually had a private tutor? Why?


Answer: Value of private tutor is not used to construct either potential outcome. Thus, both potential outcomes are independent, and `mathematics` itself is not. 

Put another way, `mathematics_1` depends on our stepping in to set `private_tutor` 
to 1, regardless of the value it actually takes (or has perhaps already taken), plus
some idiosyncratic information, summarized by `mathematics_e`, whereas 
`mathematics` depends on that same idiosyncratic information, plus the observed value of
`private_tutor`, regardless of how it got to have that value.

```{r}
cor(select(d, mathematics_0, mathematics_1, mathematics, private_tutor))
```



## Confounding time

Now let's *confound* the relationship between `tutoring` and `mathematics` with reading_writing. It 
will go like this:
```
reading_writing -> mathematics (negatively)
reading_writing -> private_tutor (positively)
private_tutor -> mathematics (positively) ESTIMAND
```

```{r}
dfork <- data.frame(reading_writing = ind$reading_writing_e)
dfork <- mutate(dfork,
                private_tutor_utility = -0.5 + reading_writing * 0.5 + ind$private_tutor_e,
                private_tutor = ifelse(private_tutor_utility > 0, 1, 0),
                mathematics = 1 + private_tutor * 2 + reading_writing * (-1) + ind$mathematics_e)
```


We add our potential outcomes. These are constructed by constructing what 
each individual would have as a mathematics score if they had a private tutor
and then what they would have as a mathematics score if they did not.


```{r}
dfork <- mutate(dfork, ##           ↓
                mathematics_0 = 1 + 0 * 2 + reading_writing * (-1) + ind$mathematics_e,
                mathematics_1 = 1 + 1 * 2 + reading_writing * (-1) + ind$mathematics_e)
```
Notice, again, that these are not a function of `private_tutor`. The value of
that variable is not used to construct either potential outcome.


Question: Are potential outcomes in this case independent of whether someone actually had a private tutor?

Answer: No. They are *not* independent. 
```{r}
cor(select(dfork, mathematics_0, mathematics_1, private_tutor))
```


Now let's condition on `reading_writing`. We'll do this twice: crudely, but 
transparently by stratification, and then opaquely but efficiently with a linear model.
Specifically, we'll stratify by the variable, and then run a tiny linear model on 
the relationship between `mathematics` and `private_tutor`. We could also run a 
bivariate correlation, but we'll do it with `lm` for comparability with the 
less transparent but more efficient way.


```{r}
res_conf <- dfork %>%
  mutate(rw_stratum = cut(reading_writing, breaks = seq(-3, 3, by = .3))) %>%
  group_by(rw_stratum) %>%
  summarise(cor0 = cor(mathematics_0, private_tutor),
            cor1 = cor(mathematics_1, private_tutor),
            cor = cor(mathematics, private_tutor), 
            # fit a linear model and extract the coefficient on tutor
            lm = coef(lm(mathematics ~ private_tutor, 
                         data = cur_data()))[['private_tutor']],
            n = n()) # how many observations? sparse estimates will be noisy! 

res_conf <- na.omit(res_conf) # drop the 3-infinity upper range
res_conf
```
Code notes: `cut` is a very useful R function for binning a continuous variable
in to bins and labeling the bins. Here we have bins that are 0.3 wide from -3 to 3.

```{r}
ggplot(res_conf, aes(x = lm)) + 
  geom_histogram(fill = "#DDCBA4") + 
  geom_vline(xintercept = mean(res_conf$lm)) + 
  geom_vline(xintercept = 2, linetype = "dashed", size = 1, color = "#BA0020")
```

## Optional statistical exercise

The mean of these stratum estimates is, for mundane statistical reasons, a 
little off. The sparsity is 
controlled by how big the bins are: fuller (wider and more central) bins 
give more precise estimates, but divide the variable more crudely. smaller 
(narrower) bins give a better discrete approximation to the underlying variable
but have fewer data points in them, so the estimates are noisier, if they even
exist.

* Try increasing the stratum size to 3. Note, this is equivalent to dichotomizing
the variable into values above 0 and below 0. What happens to the average of the 
stratum estimates in relation to the right answer?

* Now try shrinking the stratum size to a value *below* 0.3. Note that sometimes
the estimates won't exist; we remove them with `na.omit` and keep on truckin.
What happens to the average of the 
strtaum estimates in relation to the right answer?

## OK, about that efficient way

Let's do this with a linear model. We'll compare
results for not controlling and controlling.
```{r}
lm(mathematics ~ private_tutor, dfork) %>% summary()

lm(mathematics ~ private_tutor + reading_writing, dfork) %>% summary()
```
And there we are. Pretty close to 2, but only if you condition on reading and 
writing.



## Daggity model

Let's recreate the model from last class

- go to daggity  (http://www.dagitty.net/dags.html#)
- input this model
Code:
dag {
bb="0,0,1,1"
E [pos="0.083,0.213"]
G [pos="0.448,0.225"]
L [outcome,pos="0.821,0.596"]
P [pos="0.810,0.261"]
W [exposure,pos="0.111,0.556"]
E -> W
G -> L
G -> W
P -> L
W -> L
}

- Add a latent variable T which affects both treatment and outcome

Question: How does the adjustment strategy change?

- Now add variable Financial Status and identify how it could affect this model. How does the adjustment strategy change with each modification?




