---
title: "Democracy and FDI"
author: "Causal inference: Lab 4"
date: "2/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

## Jensen 2003

Slide here


Refer to the slide 13 from last weeks session. 


We are replicating the Jensen study from the session. The focus of this study is to understand the effect of democracy on Foreign Direct Investment. 

In the simplified version of the analysis, we want to use this example to understand weighting in regression. 

```{r}
library(foreign)
library(tidyverse)
```

First load the data 
```{r}
jensen.cc <- read.dta("/Users/vishali/Desktop/causal inf/Session 4/Lab/aronow_samii/jensen-rep.dta")
```


The outcome measure Jensen uses here is called `Fvar5` (FDI inflows as percentage of GDP) and the causal variable of interest is `regime' (continuous variable from 1 (authoritarian) to 20 (democratic)). 

Take a quick look at regime and FDI. 

What can you infer from this?

```{r}
hist(jensen.cc$regime)
```

```{r}
hist(jensen.cc$Fvar5, 60)
```

```{r}
summary(jensen.cc$Fvar5)
```



For our purposes we won't delve into the controls. The details of all of these can 
be found in the paper. We also won't do the full elaborate analysis here, as 
the basic points about fixed effects and weighting can be made without it.

Time to fit three models. We'll construct the formulas from a vector of variable
names because we're lazy. (Actually you're not, but I am)
```{r}
X.vars <- c("var5", "market",	"lgdppc",	"gdpgrowt",	"tradeofg",
				    "overallb",	"generalg",	"country", "d2", "d3") # these are the controls

# for predicting Fvar5 from the controls and regime
mod_f <- reformulate(c("regime", X.vars), response = "Fvar5")
mod_f

# for predicting regime from the controls
regime_mod_f <- reformulate(X.vars, response = "regime")
regime_mod_f

# for predicting Fvar5 from the controls
fdi_mod_f <- reformulate(X.vars, response = "Fvar5")
fdi_mod_f
```
and fit the models
```{r}
mod <- lm(mod_f, data = jensen.cc)
mod_regime <- lm(regime_mod_f, data = jensen.cc)
mod_fdi <- lm(fdi_mod_f, data = jensen.cc)
```

```{r}
#summary(mod)
#summary(mod_regime)
#summary(mod_fdi)
```


What do the results of mod tell you about effect of regime on FDI? What can you tell about the size of the effect?

```{r}
coef(mod)
```

a quick summary of `mod` suggests that a one unit increase in regime is associated with 
0.02108 more of Fvar5.

```{r}
sd(jensen.cc$Fvar5)
```
so this would be pretty substantively small as an effect.



FWL (Frische, Waugh and Lovell) suggest that we should be able to reconstruct that via the two other models. How would you go about this? 


```{r}
res_regime <- resid(mod_regime)
res_fdi <- resid(mod_fdi)
lm(res_fdi ~ res_regime -1) # -1 means: leave out the intercept
```



How do we relate this to heterogeneous treatment effects? 

Lets go back to the weighting business. Let's make those weights and add them to the data set.

```{r}
jensen.cc <- mutate(jensen.cc, 
                    w = res_regime^2) 
```


Now, who gets the highest weights? We'll average the weights for observations
from each country in all of its observations (in different years)
```{r}
jensen.country <- summarise(group_by(jensen.cc, country),
                            mean_w = mean(w))
```

And we can look at the *least* influential countries in our ATE:
```{r}
jc <- arrange(jensen.country, mean_w)
head(jc)
```
and *most* influential:
```{r}
tail(jc)
```


If you want to get an idea bout why these are influential it's easy to plot them,
as we did in the lecture


```{r}
ggplot(filter(jensen.cc, country == "Uruguay"), 
       aes(regime, Fvar5)) + 
  geom_point() 
```
You can swap your own country names in there. Here's Germany
```{r}
ggplot(filter(jensen.cc, country == "Germany"), 
       aes(regime, Fvar5)) + 
  geom_point() 
```
and, not quite so extreme, Madagascar
```{r}
ggplot(filter(jensen.cc, country == "Madagascar"), 
       aes(regime, Fvar5)) + 
  geom_point() 
```
a sudden lurch in regime and variation in Fvar5 (in the opposite direction to the 
hypothesis, I note) means that this country's data 'matters more' to the final 
coefficient value than somewhere like Germany does.

# Part 2

Question 2:
Does controlling for "Z" in the following variables lead to bias reduction or amplification or an increase in precision?

X is the treatment and Y is the outcome

A) X <- Z -> M -> Y

B) X -> Y <- Z <- T

C) X -> R -> Y -> Z 

D) X -> Z -> R -> Y

